# Within-Participant Analysis

There are lots of situations where we might have multiple measures of the same units of analysis. We could imagine measure a car's efficiency, and then making an upgrade to the engine, and measuring its efficiency after the upgrade. Perhaps we could test a tutoring intervention by giving students a test on a topic, allowing them to get tutored, and then measuring their knowledge post-tutoring session.

For the example in this chapter, we will focus on the `ChickWeight` dataset. To keep things simple, we are going to focus on only two timepoints in the dataset. The chunk below selects the rows we want and puts the data in wide form. For each chick, we are interested in its weight at birth and on day 2 of it's life. 

```{r}
chick = ChickWeight %>% 
  filter(Time %in% 0:2) %>% 
  pivot_wider(
    names_from = 'Time',
    names_prefix = 'weight_',
    values_from = 'weight',
    id_cols = c('Chick', 'Diet')
  )
```

Here's what the first few rows of the data look like:

```{r}
chick %>% 
  head() %>% 
  flextable()
```

## Computing Difference Scores

Our primary interest is whether chicks weigh more, on average, on day 2 than on day 0. We have all the tools we need to answer this question from Chapter \@ref(regression), with one exception.

Notice that in the regression chapter, we had a single variable, MPG, as the outcome. Here, we have two variables we are dealing with: a weight on day 0, and a weight on day 2 (`weight_0` and `weight_2` respectively).

What we really want is a variable that shows the extent to which a chick weighs more on day 2 than on day 0. It is simple to calculate this variable: we simply subtract `weight_0` from `weight_2` for each chick. Take a look at the result:

```{r}
chick_2 = chick %>% 
  mutate(
    weight_diff = weight_2 - weight_0
  )

chick_2 %>% 
  head() %>% flextable()
```

We can see that the first chick weighed 9 grams more on day 2 than on day 1. The second chick also weighed 9 grams more on day 2 than 0. The third chick weighed 4 grams *less* on day 2 than on day 0. We know this because the difference is negative; that only occurs if we subtract a larger number from a smaller one.

It might seem obvious, but it is worth remembering these rules for interpreting difference scores:

- Positive numbers indicate growth, i.e., larger scores on the second variable than on the first.
- Zero indicates no change at all. The two numbers are equal.
- Negative numbers indicate shrinking, i.e., the first number is larger than the second.

Note that these rules depend on subtracting in a particular direction. Carefully choose your subraction direction based on the situation of your analysis.

## Using Regression

Let's analyze our chick weight difference scores. We use a slightly different syntax in the following model call. Notice that the model I specify is `weight_diff ~ 1`. The "1" tells R that we do not have any variables we are using as predictors, but that it should still estimate an intercept for our model.

```{r}
wt_diff_model = lm(
  weight_diff ~ 1,
  data = chick_2
)

lmSummary(wt_diff_model)
```

Let's apply our mantras for interpretting regression. We only have an intercept, so here is the interpretation:

> On average, chicks weighed 8.16 grams more on day 2 than on day 0, $b = 8.16, F(1, 49) = 253.03, p < .001$

That's it! In the case of only 2 measurements, difference scores are straightforward to analyze. We simply take the mean of a difference score (via an intercept in regression), statistically test it against zero, and interpret accordingly.

## Important Aside: Contrast Codes

Coming soon.

## Predicting a Difference Score

We do not have to be satisfied with a simple difference score analysis. We can also predict our difference score from another variable!

More coming soon.
