--- 
title: "URFP Quantitative Project Textbook"
author: "Christopher Mellinger, Ph.D."
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
description: |
  This "textbook" contains a very brief introducion to skills that are needed for the analyses the
  students planned.
---



# Introduction

This book is intended to cover *just* enough statistical content to help students in the Undergraduate Research Fellowship Program (URFP) at the Renée Crown Wellness Institute perform analyses of [UPWARD](https://cmell.github.io/UPWARDMasterMethodsBook/) data. This book covers:

- Basic data operations.
- Regression.
- Within-participant analysis of two repeated measures via difference scores and regression.
- Moderation with regression, allowing both continuous and categorical variables.
- Mediation.

Each of these topics has been well-documented in other sources. The approach here will be to provide basic examples of useful skills, but then refer the reader to other links that elaborate and provide more foundation.

## R Help

It is helpful to know about R's built-in documentation system. At any point, you can use the console to access manuals for any R function or dataset. Just type:

```{r}
?mean
```

The help page for that function appears in the help pane in R Studio.

## Using R Studio

I highly recommend that you learn to take advantage of R Studio! Read [this introduction](http://rafalab.dfci.harvard.edu/dsbook/getting-started.html#rstudio) section. You might also look at the [RStudio cheat sheet](https://posit.co/wp-content/uploads/2022/10/rstudio-ide-1.pdf).

## Packages This Book Uses

This book relies on the packages loaded in the next chunk:

```{r, echo=F}
library(tidyverse)
library(CMUtils)
```

This book additionally uses these packages, but you probably don't need them:

```{r message=F, warning=F}
# packages we don't need people install
library(flextable)
library(ggbeeswarm)
```


If you need to, you can install the tidyverse this way:

```{r eval = F}
install.packages('tidyverse')
```

The `CMUtils` package was developed by the author of this textbook and lives on github. Install it like this:

```{r, eval=F}
install.packages('devtools')
devtools::install_github('Cmell/CMUtils')
```

## Sources

This book was created with the packages above as well as bookdown [@R-CMUtils; @R-bookdown, @R-tidyverse]. The methods presented here are largely based on the model comparison approach to statistical analysis [@judd_data_2017].

```{r, include=F}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
